Well-quasi-orderings on word languages \\ main 81b14ea82bdf369ba5dbe8de29fa418756900125 \\ 2025-10-15 17:43:32 +0200

### 2025-10-15 src/appendix.tex
fix: trying my best

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S_{\mathrm{split}}$ the set of words $w \in \Sigma^*$ 
    such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Let $S = S_{\mathrm{split}}~\cup~\min_{\prefleq} L$\todo{kinda ugly, but i can't find a nicer way of phrasing it}. 
    Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.

@@

    Let us construct a sequence of words $\seqof{w_i}[i \in \Nat]$, where $w_i
    \defined w[\vec{n_i}]$ for some well-chosen indices $\vec{n_i} \in \Nat^k$. The goal
    being that 
    if $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$,
    then it can intersect at most \emph{two} iterated words,
    with an intersection that is long enough to successfully apply
    \cref{periodic-infixes:lem}.
    In order to achieve this,
    let us first define $s$ as the minimal size of a word $v_i$
    ($1 \leq i \leq k$) and $u_j$ ($1 \leq j \leq k+1$).
    Then,
    we consider $\vec{n_0} \in \Nat^k$ such that $\vec{n_0}$ has all 
    its components greater than $\factorial{s}$ and such that
    $w[\vec{n_0}]$ belongs to $L$. 
    Then, we inductively define 
    $\vec{n_{i+1}}$  as the smallest vector of numbers greater than $\vec{n_i}$,
    such that $w[\vec{n_{i+1}}]$ belongs to $L$, 
    and with $\vec{n_i}$ having all components greater than
    $2\card{w[\vec{n_i}]}$.

    Let us assume that $k \geq 2$ in the following proof for symmetry purposes,
    and argue later on that when $k = 1$ the same argument goes through.
    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$.
    Now, because of the chosen values for $\vec{n_j}$, there exists $1 \leq \ell \leq
    k-1$ such that
    one of the three following equations holds:
    \begin{itemize}
        \item $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}}$,
        \item $w[\vec{n_i}] \infleq u_{\ell}
            v_{\ell}^{n_{j,\ell}}$,
        \item $w[\vec{n_i}] \infleq
            v_{\ell}^{n_{j,\ell}} u_{\ell+1}$.
    \end{itemize}
    In all those cases, we conclude using \cref{powers-infixes:cor}
    that there exists $x,y \in \Sigma^+$ of size at most $s$, and 
    a number $1 \leq t \leq k$ such that for all $z \in \Nat$,
    $v_i^{z} \in \InfPeriodChain{x}$ for all $1 \leq i \leq t$,
    and
    $v_i^{z} \in \InfPeriodChain{y}$ for all $t < i \leq k$.
    In particular,
    $w[\vec{n_i}] \in \InfPeriodChain{x} u_{t} \InfPeriodChain{y}$.

    \todo[inline]{Aliaume: proof of concept proof} Let us now conclude that the
    same $x,y$ work for all $\vec{n} \in \Nat^k$. The reasoning is that if
    $w[\vec{n}]$ is of the form $u_1 x^p x'$ where $x' \prefleq x$, and $x$ is
    primitive (not a power of a smaller word), then all the iterated words
    $v_i^{n_i}$ must be powers of a conjugate of $x$ (cyclic shift of $x$), and
    all the $u_i$'s must be such that the concatenation of the $u_i$'s and
    $v_i$'s align properly with $x$. Since the length of $x$ is at most $\min
    \setof{\card{v_i}}{1 \leq i \leq k}$, we conclude that the $v_i$'s
    themselves are powers of a conjugate of $x$, and that the $u_i$'s are
    suffixes or prefixes of $x$. Now, we conclude that all the words
    $w[\vec{n}]$ are of the correct form. Similar reasoning applies for the
    other two cases.

>>>>

    Conversely, assume that $L$ is \kl{well-quasi-ordered} by the \kl{prefix
    relation}. Let us define $S_{\mathrm{split}}$ the set of words $w \in \Sigma^*$ 
    such that there exists
    two words $wu$ and $wv$ both in $L$ that are not comparable for the
    \kl{prefix relation}. Let $S = S_{\mathrm{split}}~\cup~\min_{\prefleq} L$
    Assume by contradiction that $S$ is infinite.
    Then, $S$ equipped with the \kl{prefix relation} is an infinite
    tree with finite branching, and therefore contains an infinite
    branch, which is by definition an \kl{antichain branch} for $L$.
    This contradicts the assumption that $L$ is \kl{well-quasi-ordered}.

@@

    Let us construct a sequence of words $\seqof{w_i}[i \in \Nat]$, where $w_i
    \defined w[\vec{n_i}]$ for some well-chosen indices $\vec{n_i} \in \Nat^k$.
    The goal being that if $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$,
    then it can intersect at most \emph{two} iterated words, with an
    intersection that is long enough to successfully apply
    \cref{periodic-gcd:lem}. In order to achieve this, let us first define $s$
    as the maximal size of a word $v_i$ ($1 \leq i \leq k$) and $u_j$ ($1 \leq
    j \leq k+1$). Then, we consider $\vec{n_0} \in \Nat^k$ such that
    $\vec{n_0}$ has all its components greater than $\factorial{s}$ and such
    that $w[\vec{n_0}]$ belongs to $L$. Then, we inductively define
    $\vec{n_{i+1}}$  as the smallest vector of numbers greater than
    $\vec{n_i}$, such that $w[\vec{n_{i+1}}]$ belongs to $L$, and with
    $\vec{n_i}$ having all components greater than $2\card{w[\vec{n_i}]}$.

    Let us assume that $k \geq 2$ in the following proof for symmetry purposes,
    and argue later on that when $k = 1$ the same argument goes through.
    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$.
    Now, because of the chosen values for $\vec{n_j}$, there exists $1 \leq \ell \leq
    k-1$ such that
    one of the three following equations holds:
    \begin{itemize}
        \item $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}}$,
        \item $w[\vec{n_i}] \infleq u_{\ell}
            v_{\ell}^{n_{j,\ell}}$,
        \item $w[\vec{n_i}] \infleq
            v_{\ell}^{n_{j,\ell}} u_{\ell+1}$.
    \end{itemize}

    In the sake of simplicity, we will only consider one of the three cases,
    namely $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1}$, the other
    two being similar. Because the lengths used in $\vec{n_i}$ are all
    sufficiently large, we know that for every $k$, $v_k^{n_{i,k}}$ is an infix
    of a $v_{\ell}^p$ for some non-zero $p$. Therefore, we can apply
    \cref{periodic-gcd:lem} to conclude that there exists a word $x$ such that
    every $v_k$ is a power of a conjugate of $x$ (a cyclic shift of $x$), and
    $v_\ell$ is a power of $x$. We can therefore rewrite $w[\vec{n_i}]$ as $u_1
    (\sigma_1(x))^{n_{i,1}} u_2 \cdots $, where $\sigma_k$ is some conjugacy
    operation (cyclic shift). Now, in order for $w[\vec{n_i}]$ to be an infix
    of $x^{p \times n_{j,\ell}} u_{\ell+1}$, we must conclude that all the
    $u_k$'s are suffixes or prefixes of $x$, and that they align properly with
    the $\sigma_k(x)$'s to form an infix of some power of $x$, except for the
    last one. In particular, $w[\vec{n_i}] \in \InfPeriodChain{x} u_{\ell+1}$,
    but also, every other choice of $\vec{n}$ will lead to a word in
    $\InfPeriodChain{x} u_{\ell+1}$, because the alignment constraints are
    stable under pumping.

    In the case of two iterated words, the reasoning is similar, distinguishing
    between the $v_i$'s that are occurring before and after the junction of the
    two iterated words.

### 2025-10-15 src/infixes.tex
fix: trying my best

\begin{lemma}
    \label{pumping-periods:lem}
    \proofref{pumping-periods:lem}
    Let $L \subseteq \Sigma^*$ be a language
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Let $k \in \Nat$, $u_1, \cdots, u_{k+1} \in \Sigma^*$,
    and $v_1, \cdots, v_{k} \in \Sigma^+$
    be such that
    $w[\vec{n}] \defined (\prod_{i = 1}^k u_i v_i^{n_i}) u_{k+1}$
    belongs to $L$
    for arbitrarily large values of $\vec{n} \in \Nat^k$.
    Then, 
    there exists $x,y \in \Sigma^+$ of size 
    at most $\min \setof{\card{v_i}}{1 \leq i \leq k}$
    such that
    for all \todo{proof does not actually show for all}
    $\vec{n} \in \Nat^{k}$
    one of the following holds:
    \begin{enumerate}
        \item $w[\vec{n}] \in u_1 \InfPeriodChain{x}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_{k+1}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_i \InfPeriodChain{y}$
            for some $1 \leq i \leq k + 1$.

>>>>

\begin{lemma}
    \label{pumping-periods:lem}
    \proofref{pumping-periods:lem}
    Let $L \subseteq \Sigma^*$ be a language
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Let $k \in \Nat$, $u_1, \cdots, u_{k+1} \in \Sigma^*$,
    and $v_1, \cdots, v_{k} \in \Sigma^+$
    be such that
    $w[\vec{n}] \defined (\prod_{i = 1}^k u_i v_i^{n_i}) u_{k+1}$
    belongs to $L$
    for arbitrarily large values of $\vec{n} \in \Nat^k$.
    Then, 
    there exists $x,y \in \Sigma^+$ of size 
    at most $\min \setof{\card{v_i}}{1 \leq i \leq k}$
    such that
    for all
    $\vec{n} \in \Nat^{k}$
    one of the following holds:
    \begin{enumerate}
        \item $w[\vec{n}] \in u_1 \InfPeriodChain{x}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_{k+1}$,
        \item $w[\vec{n}] \in \InfPeriodChain{x} u_i \InfPeriodChain{y}$
            for some $1 \leq i \leq k + 1$.

### 2025-10-15 wqo-on-words.lncs.tex
new: isolating the u.u.r words as a theorem for the intro
new: moving stuff to appendix
fix: trying my best

\newcommand{\repositoryUrl}{\url{https://github.com/AliaumeL/word-ordering-wqo}}

\begin{document}
%
\title{Well-quasi-orderings on word languages \\
       main edb3a6bc24be4a12fe4f383c75384a6dff9d1a3c \\
       2025-10-02 17:15:30 +0200
}
\titlerunning{WQO on Words}

\author{
        Nathan Lhote\inst{3}\orcidID{0000-0003-3303-5368} \and
        Aliaume Lopez\inst{2}\orcidID{0000-0002-4205-327X} \and
        Lia Schütze\inst{1}\orcidID{0000-0003-4002-5491}}
\authorrunning{N. Lhote and A. Lopez and L. Schütze}
\institute{
Max Planck Institute for Software Systems
 \and
INP Bordeaux, LaBRI, CNRS
 \and
Aix-Marseille University
}

@@

% If there are any appendices, we include them here.

>>>>

\newcommand{\repositoryUrl}{Undefined Anonymous Repository}

\begin{document}
%
\title{Well-quasi-orderings on word languages \\
       main 81b14ea82bdf369ba5dbe8de29fa418756900125 \\
       2025-10-15 17:43:32 +0200
}
\titlerunning{WQO on Words}

\author{
    Anonymized for review
}
\authorrunning{Anonymized for review}
\institute{Anonymized for review}

@@

% If there are any appendices, we include them here.
\appendix
\input{src/appendix.tex}
